---
title: "SARS-CoV-2 sequencing report"
output:
  pdf_document:
    latex_engine: pdflatex
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage[default]{opensans}
params:
  sequencing_lab: "Broad Institute"
  intro_blurb: "The Broad Institute Viral Genomics group, in partnership with the Genomics Platform and Data Sciences Platform, has been engaged in viral sequencing of COVID-19 patients since March 2020."
  assemblies_tsv: "assemblies.tsv"
  collab_ids_tsv: "collab_ids.tsv"
  date: !r Sys.Date()
  max_date: !r Sys.Date()
  min_date: "2020-01-01"
  min_unambig: 24000
---

```{r, setup, echo=FALSE, message=FALSE, warning=FALSE}
#knitr::opts_chunk$set(echo = params$printcode)

reportable_vocs <- c('B.1.1.7', 'B.1.351', 'P.1', 'B.1.427', 'B.1.429')

library(knitr)
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyverse)
library(viridis)
library(lubridate)
library(reticulate)
library(plotly)
library(urbnmapr)
library(sf)
```

```{r, load-data, echo=FALSE}
df_assemblies = read.table(params$assemblies_tsv, sep='\t', header=TRUE)
collab_ids = read.table(params$collab_ids_tsv, sep='\t', header=TRUE)[c('external_id', 'collaborator_id')]
collab_ids <- rename(collab_ids, sample = external_id)
```

```{r, transform-data, echo=FALSE}

# derived column: genome_status
df_assemblies <- mutate(df_assemblies, genome_status = case_when(
  assembly_length_unambiguous < params$min_unambig ~ "failed_sequencing",
  vadr_num_alerts > 0 ~ "failed_annotation",
  TRUE ~ "submittable"
))
df_assemblies$genome_status = factor(df_assemblies$genome_status, levels=c('failed_sequencing', 'failed_annotation', 'submittable'))

# derived columns: geo_country, geo_state, geo_locality
df_assemblies$geo_country <- sapply(str_split(df_assemblies$geo_loc_name, ": "), function(x) x[1])
df_assemblies$geo_state <- sapply(str_split(sapply(str_split(df_assemblies$geo_loc_name, ": "), function(x) x[2]), ", "), function(y) y[1])
df_assemblies$geo_locality <- sapply(str_split(sapply(str_split(df_assemblies$geo_loc_name, ": "), function(x) x[2]), ", "), function(y) y[2])

# fix dates
df_assemblies$run_date <- as.Date(df_assemblies$run_date)
df_assemblies$collection_date <- as.Date(df_assemblies$collection_date)

# filter to earliest allowed run_date (every single library, including controls, should have this field)
df_assemblies <- subset(df_assemblies, as.Date(params$max_date) >= run_date >= as.Date(params$min_date))

# fix missing data in purpose_of_sequencing
df_assemblies[is.na(df_assemblies$purpose_of_sequencing)] <- 'Missing'
# todo: replace '' with 'Missing'

# derived column: sample_age_at_runtime
df_assemblies$sample_age_at_runtime <- df_assemblies$run_date - df_assemblies$collection_date

# join column: collaborator_id
df_assemblies <- merge(df_assemblies, collab_ids, by="sample", all.x=TRUE)

# subset to good
df_good <- subset(df_assemblies, genome_status != "failed_sequencing")

# CDC epiweeks
data.frame(date=seq(as.Date("2020-01-01"), Sys.Date()+7, by="day")) %>%
    mutate(week=epiweek(date),year=year(date)) %>%
    ddply(.(year,week), mutate, epiweek_end=max(date)) ->
    week_calendar
df_collection_weeks <- merge(df_assemblies, week_calendar, by.x='collection_date', by.y='date', all.x=TRUE)
df_run_weeks <- merge(df_assemblies, week_calendar, by.x='run_date', by.y='date', all.x=TRUE)

df_run_weeks %>%
  ddply(.(epiweek_end,genome_status), summarise, n=length(sample)) %>%
  pivot_wider(names_from=genome_status, values_from=n) -> seq_by_week
seq_by_week$attempted <- seq_by_week$failed_sequencing + seq_by_week$failed_annotation + seq_by_week$submittable
seq_by_week$genomes <- seq_by_week$failed_annotation + seq_by_week$submittable
seq_by_week <- seq_by_week %>%
  subset(select = c(epiweek_end, attempted, genomes, submittable)) %>%
  rename(epiweek_ending = epiweek_end, samples_sequenced=attempted, genomes_assembled=genomes, genomes_submittable=submittable)

# latest week stats
earliest_rundate <- min(df_assemblies$run_date)
latest_rundate <- max(df_assemblies$run_date)
latest_epiweek <- max(df_run_weeks$epiweek_end)

df_run_weeks$is_voc <- df_run_weeks$pango_lineage %in% reportable_vocs
df_run_weeks %>%
  subset(epiweek_end == latest_epiweek) ->
  df_latest_week
df_latest_week %>%
  ddply(.(genome_status), summarise, n=length(sample)) %>%
  pivot_wider(names_from=genome_status, values_from=n) ->
  latest_counts
df_latest_week %>% subset(is_voc) %>% nrow -> latest_counts$voc
snapshot <- data.frame(
  stat = c('samples sequenced', 'genomes assembled', 'genomes submittable', 'Variants of Concern'),
  last_week = c(latest_counts$failed_sequencing + latest_counts$failed_annotation + latest_counts$submittable, latest_counts$failed_annotation + latest_counts$submittable, latest_counts$submittable, latest_counts$voc),
  cumulative = c(sum(seq_by_week$samples_sequenced), sum(seq_by_week$genomes_assembled), sum(seq_by_week$genomes_submittable), length(subset(df_run_weeks, is_voc)$sample)))
```

<!-- % setup header -->
\pagestyle{fancy}
\fancyhf{}
\fancyhead[R]{Summary of SARS-CoV-2 sequencing by the `r params$sequencing_lab`\\
Internal use report generated for \textbf{all samples} as of the epiweek ending \textbf{`r latest_epiweek`}}

\fancyfoot[L]{Created on \textbf{`r paste(Sys.Date())`} for epiweek ending \textbf{`r latest_epiweek`}.}
\fancyfoot[R]{p. \thepage}

\renewcommand{\footrulewidth}{0.4pt}


`r params$intro_blurb`

This report, generated on `r params$date`, summarizes our sequencing activity for **all patient samples** across all collaborators and states. The samples included in this report were sequenced between `r earliest_rundate` and `r latest_rundate`. This report is current as of the epiweek ending **`r latest_epiweek`**.


## Weekly summary

```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(snapshot, col.names = c(' &nbsp;', paste('week ending', latest_epiweek), 'cumulative total'))
```

\vspace{0.25in}

## Sequencing activity over time by source

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
df_assemblies %>%
  subset(!is.na(collected_by) & (collected_by != '')) %>%
  ggplot() +
  geom_histogram(aes(run_date, fill=collected_by)) +
  scale_fill_brewer(palette='Set3') +
  theme(legend.position='bottom', legend.direction = 'vertical')
```

\newpage

## Sequencing activity over time and space

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=3}
df_assemblies %>%
  subset(!is.na(geo_loc_name) & (geo_loc_name != '')) %>%
  ggplot() +
  geom_histogram(aes(run_date, fill=geo_loc_name)) +
  scale_fill_brewer(palette='Pastel1') +
  theme(legend.position='right')
```

\vspace{0.25in}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4.75}
df_assemblies %>%
  ddply(.(geo_state), summarise, n=length(sample)) %>%
  merge(get_urbn_map("states", sf=TRUE), by.x='geo_state', by.y='state_name', all=TRUE) %>%
  transform(centroid_x=st_coordinates(st_centroid(geometry))[,1],
         centroid_y=st_coordinates(st_centroid(geometry))[,2]) %>%
  ggplot() +
  geom_sf(aes(geometry=geometry, fill=n), fill='grey', color='white') +
  coord_sf(datum=NA) +
  geom_point(aes(x=centroid_x, y=centroid_y, size=n), alpha=0.5, pch=21, colour='black', fill='blue') +
  #geom_scatterpie(aes(x=centroid_x, y=centroid_y, group=run_epiweek)) +
  theme(legend.position='bottom',
    axis.text.x=element_blank(), axis.text.y=element_blank())
```

\newpage

# Sequencing performance over time

### Samples sequenced weekly

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=2}
ggplot(df_assemblies) + geom_histogram(aes(run_date, fill=genome_status)) +
	scale_fill_brewer(palette='Pastel1') + theme(legend.title=element_blank())
```

This describes the total number of patient samples sequenced in this data set,
plotted by the date of the sequencing run in our lab.

"Submittable" genomes pass all QC checks and are quickly released to public genome repositories.
"Failed sequencing" are samples that failed to produce at least `r sprintf("%d", params$min_unambig)` unambiguous base pairs of
viral genome. Raw data from these samples are submitted to NCBI's SRA database, but the genomes
are not used for any analyses. "Failed annotation" are samples that produced a sufficiently complete
genome, but did not pass NCBI's VADR quality checks.

### Tabular view by CDC epiweek
```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(seq_by_week, col.names = gsub("[_]", " ", names(seq_by_week)))
```

# Timeliness of surveillance

### Sequencing date vs collection date

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(df_assemblies, height=3, width=6, units='in') + geom_point(aes(
  x=collection_date, y=run_date, color=as.numeric(run_date - collection_date, units="days"))) +
  geom_abline(linetype='dashed') +
  scale_color_viridis(name='sample_age (days)', option='plasma', discrete=FALSE)
```

This plot describes the "timeliness" of the sequencing run for the purpose of real-time surveillance of circulating lineages and variants of interest. Note that this plot likely includes many samples that were sequenced for non-surveilance purposes.


# Evolutionary Clock

### Genetic distance root-to-tip vs sample collection date

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#rtt_slope <- 0.0008 * as.numeric(as.Date('2021-01-01') - as.Date('2020-01-01'))
#rtt_intercept <-  rtt_slope * as.numeric(as.Date("2019-12-26"))
ggplot(df_good, height=3, width=6, units='in') + geom_point(aes(
  x=collection_date, y=dist_to_ref_snps, color=nextclade_clade), alpha=0.8) + scale_colour_brewer(name='Nextclade Clade', palette='Paired') + geom_smooth(aes(x=collection_date, y=dist_to_ref_snps), method=lm)
```

A "root-to-tip plot" plots the genetic distance of each sample from Wuhan Hu-1 against the date it was collected. It is generally somewhat linear. Outliers on this plot may be indicative of laboratory or metadata errors, or of evolutionarily unusual lineages (such as B.1.1.7).


\newpage

# Phylogenetic Clades and Variants

### Reportable Variants of Concern (VoCs) by CDC epiweek of sample collection

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_collection_weeks %>%
  subset(genome_status != 'failed_sequencing') %>%
  subset(pango_lineage %in% reportable_vocs) %>%
  subset(!is.na(epiweek_end)) -> x
if(nrow(x) > 0) {
  x %>%
    ddply(.(epiweek_end,pango_lineage), summarise, n=length(sample)) %>%
    pivot_wider(names_from=pango_lineage, values_from=n, values_fill=0) %>%
    rename(Collection_epiweek_ending = epiweek_end) -> vocs_by_week
} else {
  vocs_by_week <- data.frame(matrix(ncol = 1, nrow = 0))
  colnames(vocs_by_week) <- 'Collection epiweek ending'
}
kable(vocs_by_week, col.names = gsub("[_]", " ", names(vocs_by_week)))
```

### Nextclade classifications vs sample collection date

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(subset(df_collection_weeks, genome_status != "failed_sequencing")) +
  geom_histogram(aes(epiweek_end, fill=nextclade_clade)) +
	scale_fill_brewer(palette='Set1')
```

This shows the breakdown of major phylogenetic clades over time, using the Nextclade naming system. Variants of Concern (VoCs) are highlighted as specially named Nextclade clades. Nextclade clade 20I/501Y.V1 corresponds to PANGO lineage B.1.1.7, 20H/501Y.V2 corresponds to B.1.351, and 20J/501Y.V3 corresponds to P.1.

