---
title: "SARS-CoV-2 sequencing report"
output:
  pdf_document:
    latex_engine: pdflatex
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage[default]{opensans}
params:
  sequencing_lab: "Example Sequencing Lab"
  intro_blurb: "Example introductory paragraph that describes how Example Sequencing Lab has been engaged in viral sequencing of COVID-19."
  assemblies_tsv: "assemblies.tsv"
  date: !r Sys.Date()
  min_unambig: 24000
  voc_list: "B.1.1.7,B.1.351,P.1,B.1.427,B.1.429,B.1.429.1"
  voi_list: "B.1.525,B.1.526,P.2"
---

```{r, setup, echo=FALSE, message=FALSE, warning=FALSE}
#knitr::opts_chunk$set(echo = params$printcode)

reportable_vocs <- unlist(strsplit(params$voc_list, ","))
reportable_vois <- unlist(strsplit(params$voi_list, ","))

library(knitr)
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyverse)
library(viridis)
library(lubridate)
library(reticulate)
library(plotly)
library(urbnmapr)
library(sf)
```

```{r, load-data, echo=FALSE}
df_assemblies = read.table(params$assemblies_tsv, sep='\t', header=TRUE)
```

```{r, transform-data, echo=FALSE}

# fix dates
df_assemblies$run_date <- as.Date(df_assemblies$run_date)
df_assemblies$collection_date <- as.Date(df_assemblies$collection_date)

# fix missing data in purpose_of_sequencing
df_assemblies[is.na(df_assemblies$purpose_of_sequencing)] <- 'Missing'

# derived column: sample_age_at_runtime
df_assemblies$sample_age_at_runtime <- df_assemblies$run_date - df_assemblies$collection_date

# subset to good
df_good <- subset(df_assemblies, genome_status != "failed_sequencing")

# CDC epiweeks
data.frame(date=seq(as.Date("2020-01-01"), Sys.Date()+7, by="day")) %>%
    mutate(week=epiweek(date),year=year(date)) %>%
    ddply(.(year,week), mutate, epiweek_end=max(date)) ->
    week_calendar
df_collection_weeks <- merge(df_assemblies, week_calendar, by.x='collection_date', by.y='date', all.x=TRUE)
df_run_weeks <- merge(df_assemblies, week_calendar, by.x='run_date', by.y='date', all.x=TRUE)

df_run_weeks %>%
  ddply(.(epiweek_end,genome_status), .drop=FALSE, summarise, n=length(sample)) %>%
  pivot_wider(names_from=genome_status, values_from=n) -> seq_by_week
seq_by_week$attempted <- seq_by_week$failed_sequencing + seq_by_week$failed_annotation + seq_by_week$submittable
seq_by_week$genomes <- seq_by_week$failed_annotation + seq_by_week$submittable
seq_by_week <- seq_by_week %>%
  subset(select = c(epiweek_end, attempted, genomes, submittable)) %>%
  rename(epiweek_ending = epiweek_end, samples_sequenced=attempted, genomes_assembled=genomes, genomes_submittable=submittable)

# latest week stats
earliest_rundate <- min(df_assemblies$run_date)
latest_rundate <- max(df_assemblies$run_date)
latest_epiweek <- max(df_run_weeks$epiweek_end)

df_run_weeks$is_voc <- df_run_weeks$pango_lineage %in% reportable_vocs
df_run_weeks$is_voi <- df_run_weeks$pango_lineage %in% reportable_vois
df_run_weeks %>%
  subset(epiweek_end == latest_epiweek) ->
  df_latest_week
df_latest_week %>%
  ddply(.(genome_status), .drop=FALSE, summarise, n=length(sample)) %>%
  pivot_wider(names_from=genome_status, values_from=n) ->
  latest_counts
df_latest_week %>%
  subset(genome_status != "failed_sequencing") %>%
  subset(is_voc) %>%
  nrow -> latest_counts$voc
df_latest_week %>%
  subset(genome_status != "failed_sequencing") %>%
  subset(is_voi) %>%
  nrow -> latest_counts$voi
df_run_weeks %>%
  subset(genome_status != "failed_sequencing") %>%
  subset(is_voc) %>%
  nrow -> cumulative_counts_voc
df_run_weeks %>%
  subset(genome_status != "failed_sequencing") %>%
  subset(is_voi) %>%
  nrow -> cumulative_counts_voi

snapshot <- data.frame(
  stat = c('samples sequenced', 'genomes assembled', 'genomes submittable', 'Variants of Interest (VoIs)', 'Variants of Concern (VoCs)'),
  last_week = c(latest_counts$failed_sequencing + latest_counts$failed_annotation + latest_counts$submittable, latest_counts$failed_annotation + latest_counts$submittable, latest_counts$submittable, latest_counts$voi, latest_counts$voc),
  cumulative = c(sum(seq_by_week$samples_sequenced), sum(seq_by_week$genomes_assembled), sum(seq_by_week$genomes_submittable), cumulative_counts_voi, cumulative_counts_voc)
```

<!-- % setup header -->
\pagestyle{fancy}
\fancyhf{}
\fancyhead[R]{Summary of SARS-CoV-2 sequencing by the `r params$sequencing_lab`\\
Internal use report generated for \textbf{all samples} as of the epiweek ending \textbf{`r latest_epiweek`}}

\fancyfoot[L]{Created on \textbf{`r paste(Sys.Date())`} for epiweek ending \textbf{`r latest_epiweek`}.}
\fancyfoot[R]{p. \thepage}

\renewcommand{\footrulewidth}{0.4pt}


`r params$intro_blurb`

This report, generated on `r params$date`, summarizes our sequencing activity for **all patient samples** across all collaborators and states. The samples included in this report were sequenced between `r earliest_rundate` and `r latest_rundate`. This report is current as of the epiweek ending **`r latest_epiweek`**.


## Weekly summary

```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(snapshot, col.names = c(' &nbsp;', paste('week ending', latest_epiweek), 'cumulative total'))
```

CDC definitions of Variants of Concern (VoCs) and Variants of Interest (VoIs) are available at: \texttt{https://www.cdc.gov/coronavirus/2019-ncov/cases-updates/variant-surveillance/variant-info.html}

\vspace{0.25in}

## Sequencing activity over time by source

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
df_assemblies %>%
  subset(!is.na(collected_by) & (collected_by != '')) %>%
  ggplot() +
  geom_histogram(aes(run_date, fill=collected_by)) +
  scale_fill_brewer(palette='Set3') +
  theme(legend.position='bottom', legend.direction = 'vertical')
```

\newpage

## Sequencing activity over time and space

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=3}
df_assemblies %>%
  subset(!is.na(geo_loc_name) & (geo_loc_name != '')) %>%
  ggplot() +
  geom_histogram(aes(run_date, fill=geo_loc_name)) +
  scale_fill_brewer(palette='Pastel1') +
  theme(legend.position='right')
```

\vspace{0.25in}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4.75}
df_assemblies %>%
  ddply(.(geo_state), summarise, n=length(sample)) %>%
  merge(get_urbn_map("states", sf=TRUE), by.x='geo_state', by.y='state_name', all=TRUE) %>%
  transform(centroid_x=st_coordinates(st_centroid(geometry))[,1],
         centroid_y=st_coordinates(st_centroid(geometry))[,2]) %>%
  ggplot() +
  geom_sf(aes(geometry=geometry, fill=n), fill='grey', color='white') +
  coord_sf(datum=NA) +
  geom_point(aes(x=centroid_x, y=centroid_y, size=n), alpha=0.5, pch=21, colour='black', fill='blue') +
  #geom_scatterpie(aes(x=centroid_x, y=centroid_y, group=run_epiweek)) +
  theme(legend.position='bottom',
    axis.text.x=element_blank(), axis.text.y=element_blank())
```

\newpage

# Sequencing performance over time

### Samples sequenced weekly

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=2}
ggplot(df_assemblies) + geom_histogram(aes(run_date, fill=genome_status)) +
	scale_fill_brewer(palette='Pastel1') + theme(legend.title=element_blank())
```

This describes the total number of patient samples sequenced in this data set,
plotted by the date of the sequencing run in our lab.

\textbf{Submittable} genomes pass all QC checks and are quickly released to public genome repositories.
\textbf{Failed sequencing} are samples that failed to produce at least `r sprintf("%d", params$min_unambig)` unambiguous base pairs of viral genome. Raw data from these samples are submitted to NCBI's SRA database, but the genomes are not used for any analyses. 
\textbf{Failed annotation} are samples that produced a sufficiently complete genome, but did not pass NCBI's VADR quality checks.

### Tabular view by CDC epiweek
```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(seq_by_week, col.names = gsub("[_]", " ", names(seq_by_week)))
```

# Timeliness of surveillance

### Sequencing date vs collection date

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(df_assemblies, height=3, width=6, units='in') + geom_point(aes(
  x=collection_date, y=run_date, color=as.numeric(run_date - collection_date, units="days"))) +
  geom_abline(linetype='dashed') +
  scale_color_viridis(name='sample_age (days)', option='plasma', discrete=FALSE)
```

This plot describes the "timeliness" of the sequencing run for the purpose of real-time surveillance of circulating lineages and variants of interest. Note that this plot likely includes many samples that were sequenced for non-surveilance purposes.


# Evolutionary Clock

### Genetic distance root-to-tip vs sample collection date

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#rtt_slope <- 0.0008 * as.numeric(as.Date('2021-01-01') - as.Date('2020-01-01'))
#rtt_intercept <-  rtt_slope * as.numeric(as.Date("2019-12-26"))
ggplot(df_good, height=3, width=6, units='in') + geom_point(aes(
  x=collection_date, y=dist_to_ref_snps, color=nextclade_clade), alpha=0.8) + scale_colour_brewer(name='Nextclade Clade', palette='Paired') + geom_smooth(aes(x=collection_date, y=dist_to_ref_snps), method=lm)
```

A "root-to-tip plot" plots the genetic distance of each sample from Wuhan Hu-1 against the date it was collected. It is generally somewhat linear. Outliers on this plot may be indicative of laboratory or metadata errors, or of evolutionarily unusual lineages (such as B.1.1.7).


\newpage

# Phylogenetic Clades and Variants

### Reportable Variants of Concern (VoCs) by CDC epiweek of sample collection

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_collection_weeks %>%
  subset(genome_status != 'failed_sequencing') %>%
  subset(pango_lineage %in% reportable_vocs) %>%
  subset(!is.na(epiweek_end)) -> x
if(nrow(x) > 0) {
  x %>%
    ddply(.(epiweek_end,pango_lineage), .drop=FALSE, summarise, n=length(sample)) %>%
    pivot_wider(names_from=pango_lineage, values_from=n, values_fill=0, names_sort=TRUE) %>%
    rename(Collection_epiweek_ending = epiweek_end) -> vocs_by_week
} else {
  vocs_by_week <- data.frame(matrix(ncol = 1, nrow = 0))
  colnames(vocs_by_week) <- 'Collection epiweek ending'
}
kable(vocs_by_week, col.names = gsub("[_]", " ", names(vocs_by_week)))
```

### Variants of Interest (VoIs) by CDC epiweek of sample collection

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_collection_weeks %>%
  subset(genome_status != 'failed_sequencing') %>%
  subset(pango_lineage %in% reportable_vois) %>%
  subset(!is.na(epiweek_end)) -> x
if(nrow(x) > 0) {
  x %>%
    ddply(.(epiweek_end,pango_lineage), .drop=FALSE, summarise, n=length(sample)) %>%
    pivot_wider(names_from=pango_lineage, values_from=n, values_fill=0, names_sort=TRUE) %>%
    rename(Collection_epiweek_ending = epiweek_end) -> vois_by_week
} else {
  vois_by_week <- data.frame(matrix(ncol = 1, nrow = 0))
  colnames(vois_by_week) <- 'Collection epiweek ending'
}
kable(vois_by_week, col.names = gsub("[_]", " ", names(vois_by_week)))
```

CDC definitions of Variants of Concern (VoCs) and Variants of Interest (VoIs) are available at: \texttt{https://www.cdc.gov/coronavirus/2019-ncov/cases-updates/variant-surveillance/variant-info.html}

### Nextclade classifications vs sample collection date

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=3}
ggplot(subset(df_collection_weeks, genome_status != "failed_sequencing"), height=3, width=6, units='in') +
  geom_histogram(aes(epiweek_end, fill=nextclade_clade)) +
	scale_fill_brewer(palette='Set3')
```

This shows the breakdown of major phylogenetic clades over time, using the Nextclade naming system. Variants of Concern (VoCs) are highlighted as specially named Nextclade clades. Nextclade clade 20I/501Y.V1 corresponds to PANGO lineage B.1.1.7, 20H/501Y.V2 corresponds to B.1.351, and 20J/501Y.V3 corresponds to P.1. At this time, no other VoC or VoI lineages have dedicated Nextclade definitions.
